---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# survey.tm


A collection of functions that significantly eases and robustifies the management and updating of translations in multilingual surveys using CAPI software, even when questionnaires undergo changes or have multiple translations.

The package automatically extracts text items from survey questionnaires, consolidates them, and cross-references them with an existing translation database hosted on Google Sheets. This process ensures that translations stay up-to-date and synchronized with the latest version of the questionnaire(s), thereby reducing manual work, simplifying the translation management, minimizing mistakes, and enhancing consistency.

The 'Translation Database', which is a Google Sheet (see example [here](https://docs.google.com/spreadsheets/d/1qnLflfVEpCm2sRIIv2NXrBRZikkiUp7JRW0zujChq5E/edit?usp=sharing)), serves as the central hub where translators can add translations to text items of respective questionnaires. It offers a simple and intuitive user interface that's good not only for managing the translation tool but also for translators. It's easily scalable to multiple languages and questionnaires, enabling translators to track their progress, flag text items for future review or revision, and making the translation of similar types of surveys faster and more consistent. With a setup time of just 5 minutes, this package is designed to save you hours in translation management.

Please note: Currently, the package is focused on the [Survey Solutions](https://mysurvey.solutions/en/) software system, with plans to extend support to ODK-based software in the future.

## Prerequisites

- R installed on your machine. If not, download and install R from [here](https://cran.r-project.org/)
- A valid Google account for accessing [Google Sheets](https://www.google.com/sheets/about/).
- A CAPI questionnaire within [Survey Solutions](https://designer.mysurvey.solutions/questionnaire/my).

Once you have the prerequisites ready, proceed with the installation of `survey.tm` as outlined below.


## Installation

You can install the development version of survey.tm through

``` r
# If you don't have `devtools` installed, uncomment the line below to install it:
# install.packages("devtools")
devtools::install_github("RowSquared/survey.tm", ref = "main")
```


## Set up


To prepare for working with one or more questionnaires in a data collection project, you have to set up your "Translation Database Sheet". Typically, one "Translation Database Sheet" is recommended per project, even if the project involves multiple questionnaires and languages. You have two options to set it up: 


### Option 1: Use `setup_tdb()` Function

Run the `setup_tdb()` function **once**(!) to generate a Google worksheet that will act as the "Translation Database Sheet". 


Note: If you haven't used the `googlesheets4` package before, you'll need to generate an access token. Follow the instructions in the console and browser after running `setup_tdb()`.

```  {r setup, eval=FALSE}
# Optional: Indicate to `googlesheets4` which email you will use. This avoids selecting a pre-authorised account manually.
# googlesheets4::gs4_auth(email = "your.email@address.com")


# Run the code below once(!). Comment out (or remove) it afterwards.
# setup_tdb(
#   ssheet_name = "Project X: Translation Sheet", # Name to assign to new google worksheet
#   lang_names = c("German", "French") # Translations/Languages needed. Needs to be ISO 639-1 language name.
# )
```


### Option 2: Use a Template Sheet

Alternatively, you can [create a copy of this template sheet](https://docs.google.com/spreadsheets/d/1xLHEDm5bgtv4IHHCbfyMf_sdNsdzz_AUcRrnpb2fXTw) and manually set it up:

1. **Rename**: After copying, rename the existing sheet to correspond with the language you're working on (e.g., "German").
2. **Add Sheets**: If you're working with multiple languages, add more sheets as needed and name them accordingly.




## Basic Workflow

To get started with the essential features of the 'Translation Management', follow the basic steps outlined below, which will guide you through the necessary steps for a smooth ride. 


### Step 1: <span style="font-weight: normal"> Pull all data sources</span>

Pull the Questionnaire Template(s) and current 'Translation Database', compare and produce an updated translation database object. 

#### 1.1 Pull the Questionnaire Template(s) 

<details>
<summary>
Click to see detailed description
</summary>

To retrieve the current version and text items of the questionnaire(s) you're working with, you can use the `get_suso_tfiles()` function. This function downloads the [questionnaire templates](https://docs.mysurvey.solutions/questionnaire-designer/toolbar/multilingual-questionnaires/) in Excel format from the Survey Solutions Designer and stores them as nested lists in your environment. You can specify the questionnaires by using their questionnaire ID, which is a 32-character alphanumeric identifier. 

You can find this ID by logging in to the Survey Solutions Designer and accessing your questionnaire. The URL in your browser should look like `https://designer.mysurvey.solutions/questionnaire/details/xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx`, where the combination of x's after `/details/` represents your questionnaire ID.

</details>


```  {r get_suso_tfiles, eval=FALSE}
#One needs to pull the Questionnaire Template(s) ('Translation File') from the Survey Solutions Designer

# Define the questionnaire IDs you want to retrieve translations for.
questionnaires <- c("9e37f1c59e1c47039167928481cf42d2", "702bef443dfb4fc8bd8c754612590875")

# Retrieve the 'Source Questionnaire' template files
suso_trans_templates <- get_suso_tfiles(
  questionnaires = questionnaires,
  user = "your_email@example.com", # Registered with Survey Solutions Designer
  password = "your_password"
  )

```


#### 1.2 Parse Questionnaire Templates

<details>
<summary>
Click to see detailed description
</summary>

Text items to be translated are currently distributed across questionnaires and sheets in the object returned by `get_suso_tfiles()`. To aggregate all these items into a single consolidated data.table, use the `parse_suso_titems()` function. 

</details>


```  {r parse_suso_titems, eval=FALSE}
#Aggregate and consolidate text items from multiple questionnaires and sheets into a single data.table using parse_suso_titems().
source_titems <- parse_suso_titems(
  tmpl_list = suso_trans_templates, #Nested list as returned by `get_suso_tfiles()`
  collapse = TRUE, # Keep only unique text items across questionnaires
  qcode_pattern = "^Q\\d+.\\s" #Remove question coding (e.g. Q1.). Needs to be added later in `create_suso_file`
)
```  


#### 1.3 Pull 'Translation Database' Google Sheet data

<details>
<summary>
Click to see detailed description
</summary>

In order to add translations to the 'Source Questionnaires', it's necessary to pull any existing translations (if available) specified by the Translator from the 'Translation Database' Google Sheet.

`get_tdb_data()` is a simple wrapper for `googlesheets4::read_sheet()` and returns a list where each element is a sheet in the Google Sheet, representing translation data for a specific language.

Please note: At first run after `setup_tdb()`, all elements of the list returned will be empty data.table's. 
</details>

```  {r get_tdb_data, eval=FALSE}
# googlesheets4::gs4_auth(email = "your.email@address.com")

#Pull all columns and rows found in all sheets in 'Translation Database' Google Sheet
tdb_data <- get_tdb_data(ss = "GOOGLE-SHEET-IDENTIFIER")

```  


#### 1.4 Update 'Translation Database' object

<details>
<summary>
Click to see detailed description
</summary>

Compares 'Source Questionnaire' text items (object returned by `parse_odk_titems()` or `parse_suso_titems()`) against the 'Translation Database' list returned by `get_tdb_data()`.

Adds any new text item from source questionnaire(s) not yet found in the database. Text items in any sheet in the translation database object that no longer is part of the source questionnaire(s) does remain in the object but status is changed to `outdated`.

List returned will be used for subsequent processes, including creating 'Translated Questionnaires' for upload to CAPI system and/or pushed to the 'Translation Database'

</details>

```  {r update_tdb, eval=FALSE}
#Update the 'Translation Database' object based on the text items in current version of 'Source Questionnaire(s)'
new_tdb <- update_tdb(
  tdb = tdb_data,
  source_titems = source_titems
)

```  

### Step 2: <span style="font-weight: normal"> Generate outputs for CAPI program and update the Translation Database</span>


#### 2.1 Push new 'Translation Database' to Google Sheet


<details>
<summary>
Click to see detailed description
</summary>

Object `new_tdb` contains the most recent text items from the source questionnaire along with the previously added translations (if any) for all languages in the 'Translation Database'. As a next step, one should update the Google Worksheet 'Database' so that Translators can continue working on the most recent version of the questionnaire(s). To this end, use `write_tdb_data()` which writes one language from `new_tdb` object to one particular sheet in the Google Worksheet.  

Please note: It is advisable to coordinate with the Translator(s) to not overwrite translations that she is adding to the Database sheet while you are about to push new ones. 

</details>

```  {r write_tdb_data, eval=FALSE}
#Write one particular (updated) language to the 'Translation Database' Google Sheet
write_tdb_data(new_tdb[["German"]],
    ss = "GOOGLE-SHEET-IDENTIFIER",
    sheet = "German"
  )

```  


#### 2.2 Create 'Translation File' for upload to CAPI software


<details>
<summary>
Click to see detailed description
</summary>

Using object `new_tdb`, one can now add the most recent translation to a questionnaire source template (as returned by `get_suso_tfiles()`) of your choice. 

To this end, use `create_suso_file()` which will write a .xlsx file that one can use to upload to the Survey Solutions Designer. 

</details>

```  {r create_suso_file, eval=FALSE}
# Take the 'German' Translation from database and merge into the source questionnaire as returned by `get_suso_tfiles()`}
# Consider only text items of particular statuses defined by Translator
#Attention: If you used parameter qcode_pattern at `parse_suso_titems()`, you should specify here again to ensure strings are merged correctly
create_suso_file(
  tdb.language = new_tdb[["German"]],
  source_questionnaire = suso_trans_templates[["NAME-OF-QUESTIONNAIRE"]],
  statuses = c("machine", "reviewed", "translated"),
  qcode_pattern = "^Q\\d+.\\s",
  path = "your-path/German_NAME-OF-QUESTIONNAIRE.xlsx"
)

# Or loop through all source questionnaires pulled via `get_suso_tfiles()`and create a file for each language sheet on 'Translation Database'
#languages <- names(new_tdb)
#questionnaires <- names(suso_trans_templates)
#purrr::walk(languages, ~{
#  lang <- .x
#  purrr::walk(questionnaires, ~create_suso_file(
#    tdb.language = new_tdb[[lang]],
#    source_questionnaire = suso_trans_templates[[.x]],
#    statuses = c("machine", "reviewed", "translated"),
#    path = paste0("your-path/", lang, "_", .x, ".xlsx")
#  ))
#})


```  



## Utility Functions
In addition to the basic workflow functions, there are a number of additional wrappers that help in translation as well as validating the translation. 

### Identifying Software-Related Mismatches in Translations

The `syntax_check` function serves to identify software-related mismatches between the 'Text_Item' and 'Translation' columns of a given translation data table. These mismatches could arise due to:

1. **HTML-tag issues**: Mismatches in HTML tags or their count between the 'Text_Item' and 'Translation' columns.
2. **Text Substitution issues**: Discrepancies in text substitutions between the 'Text_Item' and 'Translation' columns. For instance, if a text substitution like `%rostertitle%` is present in the 'Text_Item' but missing in the 'Translation'.


``` {r syntax_check, eval=FALSE}
new_tdb <- syntax_check(
  tdb=new_tdb, # Your list of translations (usually created by `get_tdb_data()` or `update_tdb()`).
  pattern = "%[a-zA-Z0-9_]+%" # Regular expression that identifies text items in 'Text_Item' and 'Translation'. The default pattern is `%[a-zA-Z0-9_]+%`, which matches substitutions like `%rostertitle%`.
)


```

### Query Translations 

Quickly translate missing 'Translation' items using API providers:


#### Google Translate API

Simple wrapper for the Google Translate API, utilizing the [`googleLanguageR`](https://CRAN.R-project.org/package=googleLanguageR) package. For more details on setting up Google Translate API and obtaining authentication details, visit [Google Cloud Translation](https://cloud.google.com/translate/docs/setup).

```  {r query_gl_api, eval=FALSE}
# Query Google API
batchTranslate_GApi(new_tdb,
                    auth="my_path/my_service_account_key.json")
```

#### DeepL


Simple wrapper for the DeepL API, utilizing the [`deeplr`](https://github.com/zumbov2/deeplr#readme) package. For more details on setting up the API, visit [DeepL API](https://www.deepl.com/pro/change-plan#developer).


```  {r query_deepl_api, eval=FALSE}
batchTranslate_Deepl2(new_tdb,
  API_key = my_api_key
)
```




## TODO

- [ ] A main wrapper function?

### Misc.
- [ ] html-tag check: Have only those tags that are different added to comment. 
- [ ] html-tag check: check again with heavy formatting, e.g. '953faa24e13144ac984e1ad62593aab5'
- [ ] Add a sheet on translation instructions. Also stress that multiple sheets for separate language, but contain all the same (if no manual tweaks are done by user)
- [ ] `get_suso_tfiles`: print feedback to user
- [ ] Cleanup of function names & argument naming convention to make clear between translation and questionnaire
- [ ] Unit Tests
- [ ] pkgdown
- [ ] Improve similarity score function
- [ ] TODOs in functions (: 
- [ ] If new questionnaire added for which text item exist, not reflected in "QUestionnaire(s)" cell (sticks with old questionnaire)
- [ ] Remove the value.unique in Google Sheet? 
- [ ] move these todos into github issues (: 
- [ ] Color coding as dedicated function? Not really?  It is a one liner. Add to doc as an example?


### Get SuSo Designer Template 
- [ ] Confirm issue if special unicodes & readxl (e.g with "953faa24e13144ac984e1ad62593aab5")

### Long-term 
- [ ] allow local qnr template file to be read instead of pulled? 
- [ ] ODK Stream


